
% ACKNOWLEDGMENTS

\chapter*{Acknowledgments}
\thispagestyle{empty}

% Thanks to Bryan
I owe a great amount of gratitude to my advisor, Zhong Shao, for
his constant guidance, intellectual insight and extraordinary patience,
and for giving me the freedom to explore my own route.
His impact on my research, as well as my whole life, is beyond words.

% Thanks to committee members
I would like to express my sincere gratitude to my committee members,
Ruzica Piskac, Mahesh Balakrishnan, and Adam Chlipala, for many insightful
discussions on my research.
% faculty
\ignore{I also wish to thank other faculty members at Yale for their additional 
guidances and stimulating discussions on specific research topics:
Dana Angluin, James Aspnes, Ruzica Piskac, and Zhong Shao.}

This research work benefited a lot from 
the collaboration with Xiongnan(Newman) Wu,
Tahina Ramananandro,
J\'{e}r\'{e}mie Koenig,
Haozhong Zhang,
Yu Guo,
Hao Chen,
Jieung Kim,
David Costanzo,
Vilhelm Sj\"{o}berg,
and Shu-Chun Weng.
This work was improved by the valuable feedback of 
Quentin Carbonneaux,
Jan Hoffmann,
Ennan Zhai, Mengqi Liu,
and the thirty three anonymous reviewers who pored over many drafts
of this work.
% members 
I was fortunate to work alongside 
amazing colleagues of the Flint group at Yale.\ignore{,
including Henry Corrigan-Gibbs, Michael F. Nowlan, Liang Gu, 
Daniel Jackowitz, John Maheswaran, Ewa Syta and Weiyi Wu.} 
I enjoyed every minute with them.

% Family
Last but not least, I want to thank my parents 
for their unending encouragement and unwavering support, 
and thank my wife, Hao Mei,
for her love, support and understanding.

\clearpage
